---
title: "Making Maps with R"
author: "Shaun Truelove"
date: "Friday, June 19, 2015"
output: ioslides_presentation
fontsize: 10pt
---

## Making Maps with R

- rworldmap
- ggmaps

## Setup your packages

```{r Install, eval=FALSE}
library(rworldmap)
library(rworldxtra)
library(RColorBrewer)
library(maptools)
```

```{r, echo=FALSE}
# setwd("C:/Users/Shaun/SkyDrive/JHSPH/R Class/Mapping Module")
library(rworldmap)
library(rworldxtra)
library(RColorBrewer)
library(maptools)
```

## Load the map data

First we need to load the data. For this we will be using data provided by the package `rworldmap` for our maps. You can do the same thing with your own map files or using packages that use freely available maps, like `Google Maps` or `OpenStreetMap` (https://www.openstreetmap.org).

```{r getMap}
worldmap <- getMap(resolution = "high")
dim(worldmap)
```

## What's in the data object
```{r worldmapdetails}
names(worldmap)
```

## Plot the world

```{r}
par(mar=c(0,0,0,0))     # Set 0 margins
plot(worldmap)          # Plot
```

## Plot a smaller area
Setting the `xlim`  and `ylim` sets our plotted area to a specific limit of lattitude and longitude coordinates.   

- ylim = Lattitude
- xlim = Longitude

```{r PlotEurope1, fig.height=3, fig.width=4}
par(mar=c(0,0,0,0))     # Set 0 margins
plot(worldmap, xlim = c(-20, 59), ylim = c(35, 71), asp = 1) 
```

## Plot a Region, Country, or other Area
We can also select only certain regions or countries if we want.

```{r}
t(t(table(worldmap$REGION)))
```
## More 
```{r}
table(worldmap$GEO3)
```

## Plot Europe
```{r PlotEurope2, fig.height=3, fig.width=4}
par(mar=c(0,0,0,0))     # Set 0 margins
europe <- worldmap[which(worldmap$REGION=="Europe"),]               
plot(europe, col="lightgreen", bg="lightblue")
```
    
This looks a little weird because of Russia and the inclusion of the island territories.

## Europe excluding Russia and Territories
```{r PlotEurope3, fig.height=3, fig.width=4}
par(mar=c(0,0,0,0))  
europe <- worldmap[which(grepl("Europe", worldmap$GEO3) & 
                             as.character(worldmap$NAME) != "Russia"),]               
plot(europe, col="lightgreen", bg="lightblue")
```

## Zoom to an area
```{r PlotEurope4, fig.height=3, fig.width=4}
par(mar=c(0,0,0,0)) 
plot(europe, col="lightgreen", bg="lightblue", 
     xlim = c(-25, 50), ylim = c(35, 71), asp = 1)
```


Mapping Data
===========================================

## Adding Population data
First we need to load our population data:
```{r AddPop}
# setwd("C:/Users/Shaun/SkyDrive/JHSPH/R Class/Mapping Module")
world.pop <- read.csv("../data/world.population.csv",
                  header=TRUE)
row.names(world.pop) <- world.pop[,1]
```

## Check Matching
Check which countries we do not have population data for:
```{r}
country.codes <- as.character(worldmap$ADM0_A3)
worldmap$ADMIN[which(!(country.codes %in%
    world.pop$CountryCode))]
```

## Check Matching
Look for one that didn't match
```{r}
grep("west bank", world.pop$CountryName, ignore.case=TRUE, value=TRUE)
```

## Check Matching
Check what from the population data is not in the map data
```{r}
as.character(world.pop$CountryName)[
  which(!(world.pop$CountryCode %in% country.codes))]
```
  
Looks like all of the unmatched are aggregates, with a couple exceptions.

## Add the population data to the map data
```{r AddPop4}
Pop2013 <- world.pop[,c("CountryCode", "X2013")]
colnames(Pop2013)
colnames(Pop2013)[2] <- "Pop2013"

worldmap$ADM0_A3 <- as.character(worldmap$ADM0_A3)
worldmap <- merge(worldmap, Pop2013, 
  by.x="ADM0_A3", by.y="CountryCode", all.x=TRUE)
```


## Other options for adding data to map data

```{r OtherOptAdd, eval=FALSE}
joinCountryData2Map()
```
  
Part of the `rworldmap` package
  
Joins user data referenced by country codes or names to an internal map, ready for plotting using mapCountryData. Reports join successes and failures.


## Setting up to Plot
To plot population, we will give the country a color based on the population. To do this, we need to create population categories/intervals.

*Option 1.*
```{r SettingUpPopPlot, warnings=TRUE}
quantile(worldmap$Pop2013, na.rm=TRUE)
library(classInt)
brks<-classIntervals(worldmap$Pop2013[
    which(!is.na(worldmap$Pop2013))],
       n=10, style="quantile")
brks<- brks$brks
colors <- brewer.pal(length(brks), "RdYlBu")
```

## Option 2.
```{r SettingUpPopPlot2}
pop_cuts <- c(100000, 500000, 1000000, 5000000, 25000000, 
              100000000, 500000000, 1000000000, 1500000000)
colors2 <- brewer.pal(length(pop_cuts)+1, "RdYlBu")
```

## Plot the world map
```{r}
#plot the map
plot(worldmap, 
     col=colors[findInterval(worldmap$Pop2013, 
          brks, all.inside=TRUE)], 
     axes=FALSE, bg="lightgray")
```

This does not let us distinguish very well. We know the U.S. has a population of ~330 million, yet it is the same color as India and China, each with over 1 billion. We can define our population cuts better for this.

## Plot World Population - 2
```{r}
plot(worldmap, 
     col=colors2[findInterval(worldmap$Pop2013, 
      pop_cuts, all.inside=TRUE)], axes=FALSE, bg="lightgray")
title("Population by Country, World 2013") #add a title
legend("bottomleft", legend=leglabs(round(pop_cuts)), #add a legend
       fill=colors2, bty="n", cex=.6)
```


```{r, eval=FALSE}

################################################################################################
# Birth Rates

birthrates <- read.csv("../data/birthrates.csv", header=TRUE)
row.names(birthrates) <- birthrates[,1]

europe.birth <- birthrates[as.character(birthrates$CountryCode) %in% country.codes,]
europe.birth$CountryCode <- as.character(europe.birth$CountryCode)

country.codes[which(!(country.codes %in% europe.birth$CountryCode))]
europe$ADMIN[which(!(country.codes %in% europe.birth$CountryCode))]
grep("taiwan", birthrates$CountryName, ignore.case=TRUE)

births.2013 <- europe.birth[,c("CountryCode", "X2013")]
colnames(births.2013)[2] <- "BirthRate2013"

europe <- merge(europe, births.2013, by.x="ADM0_A3", by.y="CountryCode", all.x=TRUE)

quantile(europe$BirthRate2013, na.rm=TRUE)



brks<-classIntervals(europe$BirthRate2013, n=10, style="quantile")
brks<- brks$brks

pop_cuts <- c(100000, 500000, 1000000, 5000000, 25000000, 100000000, 500000000, 1000000000, 1500000000)

colors <- brewer.pal(11, "RdYlBu")

#plot the map
plot(europe, col=colors[findInterval(europe$BirthRate2013, 
      brks, all.inside=TRUE)], axes=FALSE, bg="lightgray")

#add a title
title("Birth Rate by Country, europe 2013")

#add a legend
legend("bottomleft", legend=leglabs(round(brks)),
       fill=colors, bty="n", cex=.6)

```



