---
title: "Lists and functions"
subtitle: Module 10
author: "Andrew Jaffe"
date: "June 18, 2015"
output:
  ioslides_presentation:
    css: styles.css
---

```{r loadData, comment="", prompt=TRUE, echo=FALSE}
set.seed(123)
Sal <- read.csv("../data/Baltimore_City_Employee_Salaries_FY2014.csv",as.is=TRUE)
mon <- read.csv("../data/Monuments.csv",as.is=TRUE,na.strings="")
circ <- read.csv("../data/charmcitycirc_reduced.csv",as.is=TRUE)
xx=runif(1)
```

## Review of Week Thus Far

* Reading data into R {`read.table()`}
* Subsetting vectors {[ind]} and data frames {[row,col]}
* Creating logical tests for variables in your dataset
* Creating new variables
    * Binary
    * Categorical
    * Transforming, e.g. log(), exp(), sqrt()
* Summarizing variables
    * Basic statistics, e.g. mean(), sum(), sd()
    * One variable by levels of another variable: tapply()
    * Basic exploratory plots    

You should feel comfortable doing most of the above

## Data

* We will be using multiple data sets in this lecture:
    * Salary, Monument, and Circulator from OpenBaltimore: 
    https://data.baltimorecity.gov/browse?limitTo=datasets
    * Gap Minder - very interesting way of viewing longitudinal data
        * Data is here - http://www.gapminder.org/data/
    * http://spreadsheets.google.com/pub?key=rMsQHawTObBb6_U2ESjKXYw&output=xls

## Lists

* One other data type that is the most generic are `lists`.
* Can be created using list()
* Can hold vectors, strings, matrices, models, list of other list, lists upon lists!
* Can reference data using $ (if the elements are named), or using [], or [[]]

```{r makeList, comment="", prompt=TRUE}
mylist <- list(letters=c("A", "b", "c"), 
        numbers=1:3, matrix(1:25, ncol=5))
```

## List Structure
```{r Lists, comment="", prompt=TRUE}
head(mylist)
```

## List referencing
```{r Listsref1, comment="", prompt=TRUE}
mylist[1] # returns a list
mylist["letters"] # returns a list
```

## List referencing
  
```{r Listsrefvec, comment="", prompt=TRUE}  
mylist[[1]] # returns the vector 'letters'
mylist$letters # returns vector
mylist[["letters"]] # returns the vector 'letters'
```

## List referencing

You can also select multiple lists with the single brackets. 

```{r Listsref2, comment="", prompt=TRUE}
mylist[1:2] # returns a list
```

## List referencing

You can also select down several levels of a list at once

```{r Listsref3, comment="", prompt=TRUE}
mylist$letters[1]
mylist[[2]][1]
mylist[[3]][1:2,1:2]
```

## Splitting Data Frames

The `split()` function is useful for splitting `data.frame`s

"`split` divides the data in the vector x into the groups defined by `f`. The replacement forms replace values corresponding to such a division. `unsplit` reverses the effect of split."

```{r split1, comment="", prompt=TRUE}
dayList = split(circ,circ$day)
```

## Splitting Data Frames {.smaller}

Here is a good chance to introduce `lapply`, which performs a function within each list element:

```{r lapply1, comment="", prompt=TRUE}
# head(dayList)
lapply(dayList, head, n=2)
```

---

```{r lapply2, comment="", prompt=TRUE}
# head(dayList)
lapply(dayList, dim)
```

## Writing your own functions

This is a brief introduction. The syntax is:

```
functionName = function(inputs) {
< function body >
return(value)
}
```

Then you would run the 4 lines of the code, which adds it to your workspace.

## Writing your own functions

Here we will write a function that returns the second element of a vector:

```{r return2, comment="",prompt=TRUE}
return2 = function(x) {
  return(x[2])
}
return2(c(1,4,5,76))
```

## Writing your own functions

Note that your function will automatically return the last line of code run:

```{r return2a, comment="",prompt=TRUE}
return2a = function(x) {
  x[2]
}
return2a(c(1,4,5,76))
```

And if your function is really one line or evaluation, like here, you do not need the curly brackets, and you can put everything on one line:

```{r return2b, comment="",prompt=TRUE}
return2b = function(x) x[2]
return2b(c(1,4,5,76))
```

## Writing your own functions

Also note that functions can take multiple inputs. Maybe you want users to select which element to extract

```{r return2c, comment="",prompt=TRUE}
return2c = function(x,n) x[n]
return2c(c(1,4,5,76), 3)
```

## Writing a simple function

Let's write a function, `sqdif`, that:

1. takes two numbers `x` and `y` with default values of 2 and 3.
2. takes the difference
3. squares this difference
4. then returns the final value 

## Writing a simple function

```{r sqdif, comment="",prompt=TRUE}
sqdif <- function(x=2,y=3){
     (x-y)^2
}

sqdif()
sqdif(x=10,y=5)
sqdif(10,5)
```

## Writing your own functions

Try to write a function called `top()` that takes a `matrix` or `data.frame`, and returns the first `n` rows and columns, with the default value of `n=5`.

## Writing your own functions

Try to write a function called `top()` that takes a `matrix` or `data.frame`, and returns the first `n` rows and columns

```{r top, comment="",prompt=TRUE}
top = function(mat,n=5) mat[1:n,1:n]
my.mat = matrix(1:1000,nr=100) 
top(my.mat) #note that we are using the default value for n 
```

## Custom functions in `apply`

You can use any function you want in `apply` statements. For example, from our split Circulator data

```{r top2, comment="",prompt=TRUE}
lapply(dayList, top, n = 2)
```

## Custom functions in `apply` {.smaller}

You can also designate functions "on the fly"
```{r top3, comment="",prompt=TRUE}
lapply(dayList, function(x) x[1:2,1:2])
```

## Simple apply {.smaller}

`sapply()` is a user-friendly version and wrapper of `lapply` by default returning a vector, matrix, or array

```{r sapply1, comment="", prompt=TRUE}
sapply(dayList, dim)
sapply(circ, class)
```

---

```{r sapply2, comment="", prompt=TRUE}
myList = list(a=1:10, b=c(2,4,5), c = c("a","b","c"),
                d = factor(c("boy","girl","girl")))
tmp = lapply(myList,function(x) x[1])
tmp
sapply(tmp, class)
```

---

```{r sapply3, comment="", prompt=TRUE}
sapply(myList,function(x) x[1])
sapply(myList,function(x) as.character(x[1]))
```

