---
title: "Variables"
subtitle: Module 2
author: "Andrew Jaffe"
date: "June 15, 2105"
output: ioslides_presentation
---

## Getting Started

* You should have the latest version of R installed (R 3.2.0 as of 6/15/2015)!
* Open R Studio
* Files --> New --> R Script
* Save the blank R script as "day1.R" in a directory of your choosing
* Add a comment header

## Commenting in Scripts

Add a comment header to day1.R : '#' is the comment symbol

```
#################
# Title: Demo R Script
# Author: Andrew Jaffe
# Date: 6/15/2015
# Purpose: Demonstrate comments in R
###################
 
# this is a comment, nothing to the right of it gets read

# this # is still a comment - you can use many #'s as you want

# sometimes you have a really long comment, like explaining what you 
#    are doing for a step in analysis. Take it to a second line
```

## Explaining output on slides

In slides, a command (we'll also call them code or a code chunk) will look like this
```{r code, comment="", prompt=TRUE}
print("I'm code")
```

And then directly after it, will be the output of the code.  
So `print("I'm code")` is the code chunk and [1] "I'm code" is the output.

## R as a calculator

```{r calcDemo, comment="", prompt=TRUE}
2+2
2*4
2^3
```

Note, when you type your command, R inherently thinks you want to print the result. 

## R as a calculator

* The R console is a full calculator
* Try to play around with it:
    * +, -, /, * are add, subtract, multiply, and divide
    * ^ or ** is power
    * parentheses -- ( and ) -- work with order of operations 

## R as a calculator

```{r calcDemo2, comment="", prompt=TRUE}
2+(2*3)^2
(1+3)/2 + 45
```

## R as a calculator

Try evaluating the following:
* `2+2*3/4-3`
* `2*3/4*2`
* `2^4-1`

---

* You can create variables from within the R environment and from files on your computer
* R uses "=" or "<-" to assign values to a variable name
* Variable names are case-sensitive, i.e. X and x are different

```{r assign, comment="", prompt=TRUE}
x=2
x
x*4
x+2
```

## R variables

* The most comfortable and familiar class/data type for many of you will be `data.frame`
* You can think of these as essentially Excel spreadsheets with rows (usually subjects or observations) and columns (usually variables)

* Go to RStudio --> Tools --> Import Dataset --> From Web URL then paste

http://www.aejaffe.com/summerR_2015/data/Charm_City_Circulator_Ridership.csv

```{r, echo = FALSE}
Charm_City_Circulator_Ridership <- read.csv("http://www.aejaffe.com/summerR_2015/data/Charm_City_Circulator_Ridership.csv")
```

## R variables
* We can display the top of the data with `head`:

```{r df1,  comment="", prompt=TRUE}
head(Charm_City_Circulator_Ridership)
```

## R variables
* `data.frame`s are somewhat advanced objects in R; we will start with simpler objects;
* Here we introduce "1 dimensional" classes; these are often referred to as 'vectors'
* Vectors can have multiple sets of observations, but each observation has to be the same class.

```{r assignClass, comment="", prompt=TRUE}
class(x)
y = "hello world!"
print(y)
class(y)
```

## R variables

Try assigning your full name to an R variable called `name`

## R variables

Try assigning your full name to an R variable called `name`

```{r myName, comment="", prompt=TRUE}
name = "Andrew Jaffe"
name
```

## The 'combine' function

The function `c()` collects/combines/joins single R objects into a vector of R objects. It is mostly used for creating vectors of numbers, character strings, and other data types. 

```{r assign3a, comment="", prompt=TRUE}
x <- c(1,4,6,8)
x
class(x)
```

## The 'combine' function

Try assigning your first and last name as 2 separate character strings into a single vector called `name2`

## The 'combine' function

Try assigning your first and last name as 2 separate character strings into a length-2 vector called `name2`

```{r myName2, comment="", prompt=TRUE}
name2 = c("Andrew","Jaffe")
name2
```

## R variables

`length()`: Get or set the length of vectors (including lists) and factors, and of any other R object for which a method has been defined.

```{r assign3b, comment="", prompt=TRUE}
length(x)
y
length(y)
```

##  R variables

What do you expect for the length of the `name` variable? What about the `name2` variable?

What are the lengths of each?

##  R variables

What do you expect for the length of the `name` variable? What about the `name2` variable?

What are the lengths of each?

```{r myName3, comment="", prompt=TRUE}
length(name)
length(name2)
```

## R variables

You can perform functions to entire vectors of numbers very easily.

```{r assign4, comment="", prompt=TRUE}
x + 2
x * 3
x + c(1,2,3,4)
```

## R variables

But things like algebra can only be performed on numbers.

```
> name2+4
[1] Error in name2 * 4 : non-numeric argument to binary operator
```

## R variables

And save these modified vectors as a new vector.

```{r assign5, comment="", prompt=TRUE}
y = x + c(1,2,3,4)
y 
```

Note that the R object `y` is no longer "Hello  World!" - It has effectively been overwritten by assigning new data to the variable

## R variables

* You can get more attributes than just class.  The function `str` gives you the structure of the object.

```{r assign2, comment="", prompt=TRUE}
str(x)
str(y)
```
This tells you that `x` is a numeric vector and tells you the length.

## Basic Summarization

Here are some simple functions for making calculations on data.

`sum()`: takes the sum of all numeric variables in a vector

`mean()`: takes the mean of all numeric variables in a vector

`median()`: takes the median of all numeric variables in a vector

## Back to our `data.frame` example

* Let's see what the structure of our `data.frame` is:

```{r str_charm, comment="", prompt=TRUE}
str(Charm_City_Circulator_Ridership)
```

## Review
* Creating a new script
* Using R as a calculator
* Assigning values to variables
* Performing algebra on numeric variables

