---
title: "Lab 4 Review"
subtitle: Module 4
author: "Andrew Jaffe"
date: "June 15, 2105"
output:
  ioslides_presentation:
  css: styles.css
---

```{r labSetup, echo=FALSE, message=FALSE, warning = FALSE}
mon = read.csv("../data/Monuments.csv",header=TRUE,as.is=TRUE)
```

## Question 1

1. Change the name of the "Location.1" column to "location"

## Question 1 {.smaller}

Names are just an attribute of the data frame (recall `str`) that you can change to any valid character name

Valid character names are case-sensitive, contain a-z, 0-9, underscores, and periods (but cannot start with a number).

For the `data.frame` class, `colnames()` and `names()` return the same attribute.

```{r q1, comment="", prompt=TRUE}
names(mon)
names(mon)[6] = "location"
names(mon)
```

These naming rules also apply for creating R objects

## Question 2

2. How many monuments are in Baltimore (at least this collection...)?

## Question 2

There are several ways to return the number of rows of a data frame or matrix

```{r q2, comment="", prompt=TRUE}
nrow(mon)
dim(mon)
length(mon$name)
```

## Question 3

What are the:

(a) zip codes

(b) neighborhoods

(c) council districts, and

(d) police districts 

that contain monuments, and how many monuments are in each?

## Question 3
  
`unique()` returns the unique entries in a vector

```{r q3a, comment="", prompt=TRUE}
unique(mon$zipCode)
unique(mon$policeDistrict)
unique(mon$councilDistrict)
```

---

```{r q3b, comment="", prompt=TRUE}
unique(mon$neighborhood)
```

---
  
```{r q3c, comment="", prompt=TRUE}
length(unique(mon$zipCode))
length(unique(mon$policeDistrict))
length(unique(mon$councilDistrict))
length(unique(mon$neighborhood))
```

---
  
Also note that table() can work, which tabulates a specific variable (or cross-tabulates two variables)
```{r q3d, comment="", prompt=TRUE}
table(mon$zipCode)
length(table(mon$zipCode))
```

## Question 4
  
The "by hand" way is cross-tabulating the zip codes and neighborhoods,
```{r q4a, comment="", prompt=TRUE}
tab = table(mon$zipCode, mon$neighborhood)
# tab
tab[,"Downtown"]
length(unique(tab[,"Downtown"]))
```

---
  
```{r q4b, comment="", prompt=TRUE}
tt = tab[,"Downtown"]
tt
tt == 0 # which entries are equal to 0
```

---
  
```{r q4c, comment="", prompt=TRUE}
tab[,"Downtown"] !=0
sum(tab[,"Downtown"] !=0)
sum(tab[,"Johns Hopkins Homewood"] !=0)
```

---
  
We could also subset the data into neighborhoods:
  
```{r q4d, comment="", prompt=TRUE}
dt = mon[mon$neighborhood == "Downtown",]
head(mon$neighborhood == "Downtown",10)
dim(dt)
length(unique(dt$zipCode))
```

## Question 5

How many monuments (a) do and (b) do not have an exact location/address?

## Question 5

```{r q5, comment="", prompt=TRUE}
head(mon$location)
table(mon$location != "") # FALSE=DO NOT and TRUE=DO
```

## Question 6

Which: 

(a) zip code, 
(b) neighborhood, 
(c) council district, and 
(d) police district 

contains the most number of monuments?

## Question 6 {.smaller}

```{r q6a, comment="", prompt=TRUE}
tabZ = table(mon$zipCode)
head(tabZ)
max(tabZ)
tabZ[tabZ == max(tabZ)]
```

---
  
`which.max()` returns the FIRST entry/element number that contains the maximum and `which.min()` returns the FIRST entry that contains the minimum
```{r q6b, comment="", prompt=TRUE}
which.max(tabZ) # this is the element number
tabZ[which.max(tabZ)] # this is the actual maximum
```

---
  
```{r q6c, comment="", prompt=TRUE}
tabN = table(mon$neighborhood)
tabN[which.max(tabN)] 
tabC = table(mon$councilDistrict)
tabC[which.max(tabC)] 
tabP = table(mon$policeDistrict)
tabP[which.max(tabP)] 
```

## Question 7

7. Try reading in the tab-delimited `Monuments-tab.txt` file from:
		"http://biostat.jhsph.edu/~ajaffe/winterR_2015/data/Monuments-tab.txt"
  
## Question 7
  
```{r q7, comment="", prompt=TRUE}
monTab = read.delim("http://biostat.jhsph.edu/~ajaffe/winterR_2015/data/Monuments-tab.txt",
                    header=TRUE, as.is=TRUE)
identical(mon$name,monTab$name)
```



